Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Data

### 2017-11-13 article.Rnw
Improve reparametrization section

In this section we shall describe the estimation and inference methods
adopted in this paper. The two forms of COM-Poisson models are fitted by
maximization of the log-likelihood function. The log-likelihood function
based on a set of independent observation $y_i$, $i=1,2,\ldots,n$ for
COM-Poisson$_\mu$ has the following form,
\begin{equation}
  \label{eqn:ll-rcmp}
  \ell = \ell(\bm{\beta}, \phi \mid \bm{y}) =
  e^\phi \left [
    \sum_{i=1}^n y_i
    \log \left( \mu_i + \frac{e^\phi-1}{2e^\phi} \right ) -
    \sum_{i=1}^n \log(y_i!) \right ] -
  \sum_{i=1}^n \log(Z(\mu_i, \phi)),
\end{equation}
where $\mu_i = \exp(\bm{x}_i^\top\bm{\beta})$, with
$\bm{x}_i^\top = (x_{i1},\, x_{i2},\, \ldots,\, x_{ip})$ the vector of
known covariates for the $i$-th observation, and $(\bm{\beta},\, \phi)
\in \mathbb{R}^{p+1}$. The normalizing constant $Z(\mu_i, \phi)$ is
given by
\begin{equation*}
  Z(\mu_i, \phi) = \sum_{j=0}^\infty \left [ \left (
    \mu_i + \frac{e^\phi - 1}{2e^\phi} \right )^{je^\phi}
  \frac{1}{(j!)^{e^\phi}} \right ]
\end{equation*}
To compute the log-likelihood function we need the computation of an
infinite serie for each observation, which makes the computation
expensive for regions in the parameter space where the convergence of
sum $Z$ is slow.

Parameter estimation requires numerical maximization of
\autoref{eqn:ll-rcmp}. Since $\ell$ derivatives cannot be solved in
closed forms, we adopted the BFGS algorithm \cite{Nocedal1995} that uses
numerical estimate for the Hessian matrix
$\mathcal{H}(\bm{\theta})$. Standard errors for regression coefficients
are obtained based on the observed information matrix
$\mathcal{I}(\bm{\theta})$, where
$\mathcal{I}(\bm{\theta}) = -\mathcal{H}(\bm{\theta})$ numerically
calculated by BFGS algorithm. Confidence intervals for $\hat{\mu}_i$ are
obtained by using the delta method \citep[p. 89]{Pawitan2001}.

The parameter estimation for COM-Poisson model in the original
parametrization is analogous to presented for COM-Poisson$_\mu$, however
considering \autoref{eqn:ll-rcmp} in terms of $\lambda$. Even for the
standard COM-Poisson distribution, the precision parameter is taken on
the log scale, to avoid restriction of the parameter space.

Values of the maximized log-likelihood, the Akaike criterion (AIC) and
Bayesian criterion (BIC) are computed to compare the parametric
models. As a baseline model, in the applications we fitted the
quasi-Poisson model \citep{Wedderburn1974}. This approach is based on
second-moment assumption that allows more flexibility to model. In this
case the variance of the response variable is fixed by a additional
parameter $\sigma$, $\textrm{var}(Y_i)=\sigma \mu_i$. These models are
fitted in software R \citep{Rcore2017} by function \texttt{glm(...,
  family = quasipoisson)}.

>>>>

In this section we shall describe the estimation and inference to the
two forms of the COM-Poisson regression model based on the maximum
likelihood method. The log-likelihood function for a set of
independent observations $y_i$, $i=1,2,\ldots,n$ for COM-Poisson$_\mu$
from the COM-Poisson$_\mu$ distribution has the following form,
\begin{equation}
  \label{eqn:ll-rcmp}
  \ell = \ell(\bm{\beta}, \phi \mid \bm{y}) =
  e^\phi \left [
    \sum_{i=1}^n y_i
    \log \left( \mu_i + \frac{e^\phi-1}{2e^\phi} \right ) -
    \sum_{i=1}^n \log(y_i!) \right ] -
  \sum_{i=1}^n \log(Z(\mu_i, \phi)),
\end{equation}
where $\mu_i = \exp(\bm{x}_i^\top\bm{\beta})$, with
$\bm{x}_i^\top = (x_{i1},\, x_{i2},\, \ldots,\, x_{ip})$ is a vector of
known covariates for the $i$-th observation, and $(\bm{\beta},\, \phi)
\in \mathbb{R}^{p+1}$. The normalizing constant $Z(\mu_i, \phi)$ is
given by
\begin{equation}
\label{eqn:infseries}
  Z(\mu_i, \phi) = \sum_{j=0}^\infty \left [ \left (
    \mu_i + \frac{e^\phi - 1}{2e^\phi} \right )^{je^\phi}
  \frac{1}{(j!)^{e^\phi}} \right ]
\end{equation}
The evaluation of the log-likelihood function for each observation
involves the computation of the infinite series in
\autoref{eqn:infseries}.  Thus, the fitting procedure is
computational expansive for regions of the parameter space where the
convergence of the infinite sum is slow.

Parameter estimation requires numerical maximization of
\autoref{eqn:ll-rcmp}. Since derivatives of $\ell$ cannot be obtained in
closed forms, we adopted the BFGS algorithm~\cite{Nocedal1995} as
implemented in the function \texttt{optim()} for the statistical
software \texttt{R}. Standard errors for regression coefficients are
obtained based on the observed information matrix
$\mathcal{I}(\bm{\theta})$, where
$\mathcal{I}(\bm{\theta}) = -\mathcal{H}(\bm{\theta})$ (Hessian matrix)
is computed numerically. Confidence intervals for $\hat{\mu}_i$ are
obtained by using the delta method~\citep[p. 89]{Pawitan2001}.

The parameter estimation for the COM-Poisson regression model in the
original parametrization is analogous to the one presented for the
COM-Poisson$_\mu$ distribution, however, it considers
\autoref{eqn:ll-rcmp} in terms of $\lambda$. Even for the standard
COM-Poisson distribution, the dispersion parameter is taken on the log
scale to avoid numerical issues.

In the applications we fitted the quasi-Poisson model
\citep{Wedderburn1974} as a baseline model. This approach is based on
second-moment assumption that allows more flexibility to model. In this
case the variance of the response variable is fixed by a additional
parameter $\sigma$, $\textrm{var}(Y_i)=\sigma \mu_i$. These models are
fitted in software R \citep{Rcore2017} by function \texttt{glm(...,
  family = quasipoisson)}.
